---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# `ppsr` - Predictive Power Score

`ppsr` is the R implementation of [8080labs Predictive Power Score](https://github.com/8080labs/ppscore).

The Predictive Power Score is an asymmetric, data-type-agnostic score that can 
detect linear or non-linear relationships between two columns. 
The score ranges from 0 (no predictive power) to 1 (perfect predictive power). 
It can be used as an alternative to the correlation (matrix).

Read more about the (dis)advantages of the Predictive Power Score in [this blog post](https://towardsdatascience.com/rip-correlation-introducing-the-predictive-power-score-3d90808b9598)


## Installation
You can install the development version of `ppsr` using the following R code:
```{r, eval = FALSE, echo = TRUE}
# You can get the development version from GitHub:
# install.packages('devtools')
devtools::install_github('https://github.com/paulvanderlaken/ppsr')
```

## Computing PPS
There is not the one and only way to calculate the predictive power score. 
In fact, there are many possible ways to calculate a PPS that satisfies the definition mentioned before. 
You can think of the predictive power score as a framework for a family of scores.

Currently, the `ppsr` package calculates PPS by default:

* Using the decision tree implementation of the `rpart` package, wrapped by `parsnip`
* Using 0 cross-validations
* Using F1 scores to evaluate classification models
* Using the modal or random classes as a naive benchmark for classification models
* Using MAE to evaluate regression models
* Using the mean `y` value as a naive benchmark for regression models


## Usage

The `ppsr` package has three main functions that compute PPS:

* `score()` - which computes an x-y PPS
* `score_predictors()` - which computes X-y PPS
* `score_matrix()` - which computes X-Y PPS

where `x` and `y` represent an individual feature/target, 
and `X` and `Y` represent all features/targets in a given dataset.

Examples:
```{r}
score(x = iris$Sepal.Length, y = iris$Sepal.Width)
```

```{r}
score_predictors(df = iris, y = 'Species')
```

```{r}
score_matrix(df = iris)
```


## Visualizing PPS
Subsequently, there are two main functions that wrap around these computational 
functions to help you visualize your PPS using `ggplot2`:

* `visualize_predictors()` - producing a barplot of all X-y PPS
* `visualize_matrix()` - producing a heatmap of all X-Y PPS

Examples:
```{r}
visualize_predictors(df = iris, y = 'Species')
```

```{r}
visualize_matrix(df = iris)
```



## Open issues & development
PPS is a relatively young concept, and likewise the `ppsr` package is still under development. 
The current package was built in only a few hours and will likely contain bugs and/or inefficiencies.
If you have any improvements, please raise an issue. 

On the developmental agenda are currently:

* Implementation of different modeling techniques / algorithms
* Implementation of different model evaluation metrics
* Implementation of cross-validation
* Implementation of downsampling for large datasets


Note that there's also an unfinished [R implementation of the PPS package by 8080labs](https://github.com/8080labs/ppscoreR).
