---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/paulvanderlaken/ppsr/workflows/R-CMD-check/badge.svg)](https://github.com/paulvanderlaken/ppsr/actions)
 <!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/README/",
  fig.width = 7,
  fig.height = 5, 
  dpi = 300
)
```

# `ppsr` - Predictive Power Score

`ppsr` is the R implementation of [8080labs' Predictive Power Score](https://github.com/8080labs/ppscore).

The PPS is an asymmetric, data-type-agnostic score that can detect linear or non-linear relationships between two columns. The score ranges from 0 (no predictive power) to 1 (perfect predictive power). It can be used for data exploration purposes, as a complement to the correlation (matrix).

Read more about the (dis)advantages of the Predictive Power Score in [this blog post](https://towardsdatascience.com/rip-correlation-introducing-the-predictive-power-score-3d90808b9598).


## Installation
You can install the development version of `ppsr` using the following R code:
```{r, eval = FALSE, echo = TRUE}
# You can get the development version from GitHub:
# install.packages('devtools')
devtools::install_github('https://github.com/paulvanderlaken/ppsr')
```

## Computing PPS
You can think of the predictive power score as a framework for a family of scores.
There is not one single best way to calculate a predictive power score. 
In fact, there are many possible ways to calculate a PPS that satisfy the definition mentioned before. 

Currently, the `ppsr` package calculates PPS by default:

* Using the default decision tree implementation of the `rpart` package, wrapped by `parsnip`
* Using 5 cross-validations
* Using F1 scores to evaluate classification models. 
Scores are normalized relatively to a naive benchmark consisting of predicting the modal or random `y` classes
* Using MAE to evaluate regression models. 
Scores are normalized using relatively to a naive benchmark of predicting the mean `y` value


## Usage

The `ppsr` package has three main functions that compute PPS:

* `score()` - which computes an x-y PPS
* `score_predictors()` - which computes all X-y PPS
* `score_df()` - which computes all X-Y PPS
* `score_matrix()` - which computes all X-Y PPS, and shows them in a matrix

where `x` and `y` represent an individual feature/target, 
and `X` and `Y` represent all features/targets in a given dataset.

Examples:
```{r}
ppsr::score(iris, x = 'Sepal.Length', y = 'Sepal.Width')
```

```{r}
ppsr::score_predictors(df = iris, y = 'Species')
```

```{r}
ppsr::score_matrix(df = iris)
```



## Visualizing PPS
Subsequently, there are three main functions that wrap around these computational 
functions to help you visualize your PPS using `ggplot2`:

* `visualize_pps()` - producing a barplot of all X-y PPS, or a heatmap of all X-Y PPS
* `visualize_correlations()` - producing a heatmap of all X-Y correlations
* `visualize_both()` - produces two heatmaps of the PPS and correlations of all X-Y PPS side-by-side

Examples:
```{r, PPS-barplot}
# If you specify the target variable (y), you get a barplot of its predictors
ppsr::visualize_pps(df = iris, y = 'Species')
```

```{r, PPS-heatmap}
# If you do not specify `y`, you get a PPS matrix visualized as a heatmap 
ppsr::visualize_pps(df = iris)
```

```{r, correlation-heatmap}
ppsr::visualize_correlations(df = iris)
```

```{r, sbs-heatmap, fig.width=14}
ppsr::visualize_both(df = iris)
```

You can change the colors of the visualizations using the functions arguments.
There are also arguments to change the color of the text scores.

Furthermore, the functions return `ggplot2` objects, so you can easily change the theme and other settings.
```{r, custom-plot}
ppsr::visualize_pps(df = iris,
                    color_value_high = 'red', 
                    color_value_low = 'yellow',
                    color_text = 'black') +
  ggplot2::theme_classic() +
  ggplot2::labs(title = 'Add your own title', caption = 'Or a nice caption')
```


## Parallelization

The number of predictive models that need to be built to fill a PPS matrix of a dataframe increase exponentially with every column in that dataframe. 

For high dimensional datasets, it can thus takes a decent amount of time to generate the associated PPS matrix.

One way to speed matters up is to use the `ppsr::score_predictors()` function and to focus on the predictability of a single target/dependent variable of interest.

Yet, since version `0.0.1`, all `ppsr::score_*` and `pssr::visualize_*` functions now take in two arguments that facilitate parallel computing.
You can parallelize `ppsr`'s computations by setting the `do_parallel` argument to `TRUE`. If done so, a cluster will be created using the `parallel` package. By default, this cluster will use the maximum number of cores (see `parallel::detectCores()`) minus 1.

However, with the second argument -- `n_cores` -- you can manually specify the number of cores you want `ppsr` to use.

Examples:
```{r, eval = FALSE}
ppsr::score_df(df = mtcars, do_parallel = TRUE)
```

```{r, eval = FALSE}
ppsr::visualize_pps(df = iris, do_parallel = TRUE, n_cores = 3)
```



## Open issues & development
PPS is a relatively young concept, and likewise the `ppsr` package is still under development. 
If you spot any bugs or potential improvements, please raise an issue or submit a pull request. 

On the developmental agenda are currently:

* Implementation of different modeling techniques / algorithms
* Implementation of different model evaluation metrics
* Implementation of downsampling for large datasets


Note that there's also an unfinished [R implementation of the PPS package by 8080labs](https://github.com/8080labs/ppscoreR).
